\section*{Introduction}
\addcontentsline{toc}{section}{Introduction}


Let me test some stuff I want to see how this goes: erchants of Doubt shows that large Industries like to obscure issues by flooding the scientific debate with unrelated or distracting researcherchants of Doubt shows that large Industries like to obscure issues by flooding the scientific debate with unrelated or distracting researcherchants of Doubt shows that large Industries like to obscure issues by flooding the scientific debate with unrelated or distracting researcherchants of Doubt shows that large Industries like to obscure issues by flooding the scientific debate with unrelated or distracting research.

Merchants of Doubt shows that large Industries like to obscure issues by flooding the scientific debate with unrelated or distracting researcherchants of Doubt shows that large Industries like to obscure issues by flooding the scientific debate -- with unrelated or distracting -- research. Merchants of Doubt shows that large Industries like to obscure issues by flooding the scientific debate with unrelated or distracting researcherchants of Doubt shows that large Industries like to obscure issues by flooding the scientific debate with unrelated or distracting researcherchants of Doubt shows that large Industries like to obscure issues by flooding the scientific debate with unrelated or distracting researcherchants of Doubt shows that large Industries like to obscure issues by flooding the scientific debate with unrelated or distracting research

\begin{enumerate}
    \item Merchants of Doubt shows that large Industries like to obscure issues by flooding the scientific debate with unrelated or distracting research
    \item From the recent growth of the online industry and the respective growing financial power and importance in research combined with problems in the digital space like: \begin{itemize}
        \item Privacy
        \item Hatespeech
        \item Manipulation of Elections 
        \item Power of monopolies (e.g. Facebook buying Whatsapp)
        \item Addiction to online services...
    \end{itemize} One could rightfully imagine Big Tech to engage in some form of distracting activities. 
    \item Addressing this question is notoriously difficult:\begin{itemize}
        \item Big Tech does a lot of research on the frontier of the scientific debate, thus advancing science by giving scientist access to its data.
        \item Pursuing research which is in their interest is not per se bad. It only becomes objectable if it is carried out and on a large scale and coordinated, especially if Big Tech uses their market position to do so. 
        \item Determining the influence (un)conditional funding actually has on the outcome of research is difficult, also because of lacking counterfactuals. 
    \end{itemize} 
    \item Therefore, I want to focus on transparency of research funding / rules on disclosure in the different field Big Tech is engaged in research.
    \item This will be compared with the medical industry which over the years has adopted very strict regulation on the disclosure of funding.
\end{enumerate}

The idea for the structure of this case-study is as follows:
\begin{enumerate}
    \item In the Introduction I want to give some Background on patterns of distraction explored in "Merchants of Doubt"
    \item The history of the medical industry and the transparency / disclosure of research funding.
    \item To then go on to explore the transparency of research funding provided by Google
    \item Finally, I want to explore the debate around this topic in the Big Tech environment
\end{enumerate}


\subsection{A new idea how to approach this}
\begin{enumerate}
    \item Merchants of Doubt
    \item Scale of R\textit{\&}D von Google und Co. 20 bn USD Of Google vs. 300 bn USD \textsc{total} spending on Universities in the US \citep{sah_conflict_2018}
    \item Spending 2 mn per Quarter on official lobbying
    \item Case study on Medical Disclosure of Conflicts of Interest (Uni policies...)
    \item Trying to find similarities in D\textit{\&}C Science. 
    \item Anti Cases \begin{itemize}
        \item Lütge in munich
        \item \textit{Googles in-transparency -> vs. Oracle study this in detail}
        \item Google Outreach maybe one or two cases
        \item Bergkmann Klein Center 
    \end{itemize}
\end{enumerate}